A thermodynamic view of dusty protoplanetary disks

### 2017-07-14 linear.tex
tiny intro edits (this paper is organized...), remove repeated text in SI section
moved general expression of vertical shear up into general stability section

\subsubsection{Vertical shear}\label{vertshear}
The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\tepsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\tepsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p \seff}{\p z} -\frac{\p
    P}{\p z}\frac{\p \seff}{\p r} \right). 
\end{align}
Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align}\label{vshear2}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
%  -\frac{\p\epsilon}{\p z}\left[\frac{\p\ln{\rhog}}{\p r} + \frac{q}{r}\right]\right.\notag\\
 -\frac{\p\epsilon}{\p z}\frac{\p\ln{P}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. -\frac{q}{r} \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z}
  \right\}. 
\end{align}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio. The third term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient, and survives in the dust-free limit.

@@

In a future work we will perform a detailed comparison between the
simplified one-fluid and full two-fluid frameworks in calculating SI.

>>>>

\subsubsection{Vertical shear}\label{vertshear}
Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp})
gives the disk's explicit vertical shear profile:
\begin{align}\label{vshear2}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
%  -\frac{\p\epsilon}{\p z}\left[\frac{\p\ln{\rhog}}{\p r} + \frac{q}{r}\right]\right.\notag\\
 -\frac{\p\epsilon}{\p z}\frac{\p\ln{P}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. -\frac{q}{r} \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z}
  \right\}. 
\end{align}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio. The third term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient, and survives in the dust-free limit.

@@

### 2017-07-14 model.tex
moved general expression of vertical shear up into general stability section

%\subsection{Dusty buoyancy forces}
Having identified the entropy of an isothermal dusty gas, we can also
define the disk vertical buoyancy frequency $N_z$ as   
%$N_z$ of the mixture is given by 
\begin{align}
  N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p \seff}{\p z}
  =c_s^2(r) \frac{\p \ln{\rhog}}{\p z}\frac{\p \tepsilon}{\p z},
%  \quad 
 % N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p \seff}{\p z}.
%&=
  %\frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
%  z}\frac{\p\epsilon}{\p z} \\ &
  %                                =
 % \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
%Typically in a thin, smooth disk we have $|N_r|\ll |N_z|$. 
where the second equality assumes a vertically isothermal disk. A
similar definition exists for the radial buoyancy frequency
$N_r$. Typically in a thin, smooth disk we have $|N_r|\ll |N_z|$.

@@

However, disk structures obtained this way generally do not satisfy 
the steady state effective energy Eq. \ref{eff_energy} because typically 
$\mathcal{C}\neq0$.  In order to have a well-defined stability problem, we formally
consider equilibria with $\mathcal{C}\equiv 0$. Namely      
\begin{inparaenum}[1)] 
\item 
  unstratified or 2D disks with finite dust-gas drag; %(\S\ref{si});   
\item 
  perfectly-coupled dust with $\tstop=0$. %(\S\ref{results}). 
\end{inparaenum} 
Alternatively, we are assuming that $\tstop$  is sufficiently small
such that the evolution of the background disk (e.g. dust-settling) 
occur on much longer timescales than the instabilities of interest.

@@

On the other hand, Eq. \ref{dusty_solberg2} can be violated in
disks if the dust is vertically well-mixed ($\p_z\epsilon =
0$) but  radially-varying ($\p_r\epsilon\neq0$). In this case the
left-hand-side of Eq. \ref{dusty_solberg2} becomes
\begin{align}
-\frac{c_s^4}{(1+\tepsilon)^4}\left(\frac{\p\ln{\rhog}}{\p
    z}\right)^2\left(\frac{\p\epsilon}{\p r}\right)^2<0.
\end{align}
Such an isothermal dusty disk is \emph{unstable} because there is no effective vertical 
buoyancy to stabilize vertical motions ($N_z=0$). This situation is 
identical to the pure gas, adiabtic simulations of \cite{nelson13}
where the gas entropy is vertically uniform but has a radial
gradient. The authors indeed find instability. We give a numerical
example in the dusty context in \S\ref{vert_mixed}.

>>>>

%\subsection{Dusty buoyancy forces}
Having identified the entropy of an isothermal dusty gas, we can also
define the disk vertical buoyancy frequency $N_z$ as   
%$N_z$ of the mixture is given by 
\begin{align}
  N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p \seff}{\p z}
  =c_s^2(r) \frac{\p \ln{\rhog}}{\p z}\frac{\p \tepsilon}{\p z},
%  \quad 
 % N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p \seff}{\p z}.
%&=
  %\frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
%  z}\frac{\p\epsilon}{\p z} \\ &
  %                                =
 % \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
%Typically in a thin, smooth disk we have $|N_r|\ll |N_z|$. 
where the second equality assumes a vertically isothermal disk. A
similar definition exists for the radial buoyancy frequency
$N_r$. However, in thin, smooth disks we have $|N_r|\ll |N_z|$.

@@

The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\tepsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\tepsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p \seff}{\p z} -\frac{\p
    P}{\p z}\frac{\p \seff}{\p r} \right). 
\end{align}
Since $\seff$ depends on $\tepsilon$, we see that radial gradients in
the dust-to-gas ratio also drives vertical shear \citep[cf. that 
driven by $\p_z\epsilon$ usually discussed in dusty disks,  
e.g.][]{chiang08}.

However, disk structures obtained from
Eq. \ref{steady_momr}---\ref{steady_momz} generally do not satisfy  
the steady state effective energy Eq. \ref{eff_energy} because typically 
$\mathcal{C}\neq0$.  In order to have a well-defined stability problem, we formally
consider equilibria with $\mathcal{C}\equiv 0$. Namely      
\begin{inparaenum}[1)] 
\item 
  unstratified or 2D disks with finite dust-gas drag; %(\S\ref{si});   
\item 
  perfectly-coupled dust with $\tstop=0$. %(\S\ref{results}). 
\end{inparaenum} 
Alternatively, we are assuming that $\tstop$  is sufficiently small
such that the evolution of the background disk (e.g. dust-settling) 
occur on much longer timescales than the instabilities of interest.

@@

On the other hand, Eq. \ref{dusty_solberg2} can be violated in
disks if the dust is vertically well-mixed ($\p_z\epsilon =
0$) but  radially-varying ($\p_r\epsilon\neq0$). In this case the
left-hand-side of Eq. \ref{dusty_solberg2} becomes
\begin{align}
-\frac{c_s^4}{(1+\tepsilon)^4}\left(\frac{\p\ln{\rhog}}{\p
    z}\right)^2\left(\frac{\p\epsilon}{\p r}\right)^2<0.
\end{align}
Such an isothermal dusty disk is \emph{unstable} but there is no
effective vertical buoyancy to stabilize vertical motions ($N_z=0$)
that can tap into the free energy associated with  
vertical shear due to the radial gradient in the dust-to-gas ratio. 
This situation is identical to the pure gas, adiabtic simulations of \cite{nelson13}
where the gas entropy is vertically uniform but has a radial
gradient. The authors indeed find instability. We give a numerical
example in the dusty context in \S\ref{vert_mixed}.
